---
output: "github_document"
---

```{r, include=FALSE}
shieldio_badge <- function(name, text, color, hover = NULL, href = NULL, ..., 
  url = 'https://img.shields.io/badge/{{name}}-{{text}}-{{color}}.svg',
  style = 'flat-square') {
  
  .dots <- list(...)
  name  <- RCurl::curlEscape(whisker::whisker.render(name, .dots))
  text  <- RCurl::curlEscape(whisker::whisker.render(text, .dots))
  color <- whisker::whisker.render(color, .dots)
  
  if (is.numeric(c <- type.convert(color)))
    color <- switch(letters[[floor(pmax(pmin(c, 1), 0) * 6 + 1)]],
      g = 'brightgreen',
      f = 'brightgreen',
      e = 'green',
      d = 'yellowgreen',
      c = 'yellow',
      b = 'orange',
      a = 'red')
  
  color <- RCurl::curlEscape(color)  
  url <- whisker::whisker.render(paste(url, paste0('style=', style), sep = '?'))
  hover <- if (is.null(hover)) name else hover
  whisker::whisker.render('[![{{hover}}]({{url}})]({{href}})')
}
```

```{r build.badges, include=FALSE}
failed_tests = tryCatch(devtools::test(), error = function(e) NULL)
failed_tests = if (!is.null(failed_tests) && 'failed' %in% names(as.data.frame(failed_tests))) sum(failed_tests$failed) else 1
build_badge <- shieldio_badge(
  'build', '{{p}}', '{{d}}', 'Build Status', 
  p = if (failed_tests == 0) 'passing' else 'failing', 
  d = if (failed_tests == 0) 'brightgreen' else 'red')

coverage = covr::percent_coverage(covr::package_coverage()); 
coverage_badge <- shieldio_badge(
  'coverage', '{{p}}%', '{{d}}', 'Code Coverage', 
  p = round(coverage,1), 
  d = (coverage/100-0.8)/0.2)
```


# gClinBiomarker

`r build_badge` `r coverage_badge`

## Overview

gClinBiomarker is an R package that contains functions to perform baseline and longitutinal biomarker analyses 

Source code: https://github.roche.com/Rpackages/gClinBiomarker

Project page: https://pages.github.roche.com/Rpackages/gClinBiomarker/

Documentations could be found under the "article" tab of the project page. Two documents are provided:
- A user vignette demonstrates how gClinBiomarker package may help in your biomarker analysis
- An exapmle use case document which contains more detailed use cases

pdf version of these two documents can be found [here](https://github.roche.com/Rpackages/gClinBiomarker/tree/master/inst/doc).

gClinBiomarker also provide a set of Rmarkdown templates that allow you to generate biomarker analysis report by "one click". The R markdown templates can be found at [here](https://github.roche.com/lengn/gClinbiomarker_documents), along with some slide decks.

## Getting Started

### Installation

To install this package from R, use `install_github()` function from the **devtools** package

In R, type:

```{r, eval = FALSE}
## install.packages("devtools")
library(devtools)
install_github("RPackages/gClinBiomarker", host="https://github.roche.com/api/v3")
```

Note that on bce (r.bas.roche.com), the default R is from an older version. To install, type

```{r, eval = FALSE}
install_github("RPackages/gClinBiomarker", host="github.roche.com/api/v3")
```

### Getting the Documentation

Use the command `vignette(package = 'gClinBiomarker')` to view a list of
avaialble vignettes for the `gClinBiomarker` package. Then use the command
`vignette(<vignett_name>)` to view the relevant documentation.

## gClinBiomarker Analysis and Workflows

### Supported endpoint and biomarker types

#### Endpoints

* Time-to-event
* Binary
* Continuous
* Longitudinal
 
#### Biomarkers 

* Continuous
* Ordinal
* Categorical
* Longitudinal

#### Analysis

* 2-arm study: predictive and prognostic biomarker analysis
* 1-arm study: prognostic biomarker analysis
 
### Workflow

#### Step 1:
Is your biomarker population your full patient population?
Take a look at these functions:

`SummaryVars()`, `CompareKM()`, `PlotRspBar()`

#### Step 2:
Is your biomarker response a dynamic range?
Does it have a skewed distribution?
Is it correlated with clinical variables?

`PlotProperty()`, `PlotTabForestMulti()`

#### Step 3:
Is your biomarker response associated with a clinical outcome?
Is it prognostic or predictive?
Is there an optimal biomarker cutoff with a consistent trend?

`PlotTabForestBiomarker()`, `PlotSTEPP()`

#### Step 4:

Estimate the clinical benefit within a biomarker subgroup.

`PlotKM()`, `PlotLong()`, `PlotRspBar()`, `CoxTab()`, `LogRankTab()`


```{r, include = FALSE}
# ![screen shot 2017-09-28 at 1 44 01 pm](https://media.github.roche.com/user/48/files/431d398a-a453-11e7-8801-1c6915156185)
# ![screen shot 2017-09-28 at 1 44 13 pm](https://media.github.roche.com/user/48/files/47cad168-a453-11e7-85f7-deee2f7604ab)
# ![screen shot 2017-09-28 at 1 44 24 pm](https://media.github.roche.com/user/48/files/4ad58466-a453-11e7-80be-9af0c23fcedd)
# ![screen shot 2017-09-28 at 1 44 34 pm](https://media.github.roche.com/user/48/files/507dd1a2-a453-11e7-8216-442f15bd500a)
# ![screen shot 2017-09-28 at 1 44 43 pm](https://media.github.roche.com/user/48/files/53ab85e0-a453-11e7-825d-718f28d1fca3)
```
