---
title: "gClinBiomarker: Example Use Cases"
author: "Ning Leng, Alexey Pronin, Doug Kelkhoff"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Loading packages and data

```{r setup1, message=FALSE}
if (!require(gClinBiomarker)) { 
  install_github("RPackages/gClinBiomarker",  host="https://github.roche.com/api/v3")
  library(gClinBiomarker)
}

library(knitr)
library(devtools)
library(ggplot2)
data(input)
sample.data <- input
```


# Example data set

```{r}
head(sample.data)

str(sample.data)
```

# SummaryVars(): Summarize demographics variables (check demographics imbalance)

## Default

```{r}
kable(
SummaryVars(data=sample.data,trt='Arm', subgroup='BEP', var=c('Age','Sex'), 
       var.class=c("numeric","categorical"))
)
```

## Compare to non-BEP instead of ITT
```{r}
kable(
SummaryVars(data=sample.data,trt='Arm', subgroup='BEP', var=c('Age','Sex'), 
       var.class=c("numeric","categorical"), compare.subgroup=TRUE)
)
```

## Allow testing
```{r}
kable(
SummaryVars(data=sample.data,trt='Arm', subgroup='BEP', var=c('Age','Sex'), 
       var.class=c("numeric","categorical"), test.subgroup=TRUE)
)
```

## Reorder TRT/CTRL arm in display
```{r}
kable(
SummaryVars(data=sample.data,trt='Arm', subgroup='BEP', var=c('Age','Sex'), 
       var.class=c("numeric","categorical"), trt.order=c("TRT","CTRL"))
)
```


## Combine trt arms
```{r}
kable(
SummaryVars(data=sample.data, subgroup='BEP', var=c('Age','Sex'), 
       var.class=c("numeric","categorical"))
)
```

## Alternative BEP indicator

```{r}
sample.data$BEP2 <- ifelse(sample.data$BEP==1,"Yes","No")
kable(
SummaryVars(data=sample.data,trt='Arm', subgroup='BEP2', var=c('Age','Sex'), 
       var.class=c("numeric","categorical"), subgroup.indicator="Yes")
)
```




# PlotProperty(): Plot biomarker, clinical covariate property and their association

## Numeric biomarker variable. No clinical variables. Log transformation.
```{r include=TRUE,fig.width=10, fig.height=5}
PlotProperty(data=input, biomarker.var="KRAS.exprs", biomarker.class="numeric", log2=TRUE)
```

## No biomarker variable. Two numeric clinical variables. Log transformation for one varible.
```{r include=TRUE,fig.width=10, fig.height=5}
PlotProperty(data=input, biomarker.var=NULL, biomarker.class=NULL,
             var=c("Weight","Age"), var.class=c("numeric", "numeric"),
             log2=c(TRUE, FALSE), par.param = list(mfrow=c(1,2), cex.axis=1.2))
```

## Numeric biomarker variable. Two clinical categorical variables. Log transformation for numeric variable.
```{r include=TRUE, fig.width=12, fig.height=12}
PlotProperty(data=input, biomarker.var="KRAS.exprs", biomarker.class="numeric",
             var=c("ECOG", "Country"), var.class=c("categorical", "categorical"),
             log2=TRUE, par.param = list(mfrow=c(3,2), cex.axis=1.2),
             show.clinical.uni=TRUE)
```

## Numeric biomarker variable. Two clinical variables: one is categorocal, second is numeric. Log transformation for biomarker (numeric) variable.
```{r include=TRUE, fig.width=10, fig.height=12}
PlotProperty(data=input, biomarker.var="KRAS.exprs", biomarker.class="numeric",
             var=c("Sex", "Age"), var.class=c("categorical", "numeric"),
             log2=c(TRUE, FALSE, FALSE), par.param = list(mfrow=c(3,2), cex.axis=1.4),
             show.clinical.uni=TRUE)
```

## Categorical biomarker variable. Categorical clinical variable.
```{r include=TRUE, fig.width=14, fig.height=10}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var="Country", var.class="categorical", 
             par.param = list(mfrow=c(2,2), cex.axis=1.2),
             show.clinical.uni=TRUE)
```

## Categorial biomarker variable. Categorical and numeric clinical variables. No log transformation.
```{r include=TRUE, fig.width=14, fig.height=12}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var=c("Country", "Age"), var.class=c("categorical", "numeric"),
             par.param = list(mfrow=c(3,2), cex.axis=1.2),
             show.clinical.uni=TRUE)
```

## Categorial biomarker variable. Categorical and numeric clinical variables. No log transformation. Don't plot biomarker variable but plot clinical variables and association between biomarker and clinical variables.
```{r include=TRUE, fig.width=14, fig.height=10}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var=c("Country", "Age"), var.class=c("categorical", "numeric"),
             par.param = list(mfrow=c(2,2), cex.axis=1.2), show.biomarker.uni = FALSE,
             show.clinical.uni=TRUE)
```

## Categorial biomarker variable. Categorical and numeric clinical variables. No log transformation. Don't plot clinical variables but plot biomarker variable and association between biomarker and clinical variables.
```{r include=TRUE, fig.width=14, fig.height=10}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var=c("Country", "Age"), var.class=c("categorical", "numeric"),
             par.param = list(mfrow=c(2,2), cex.axis=1.2))
```

## Categorial biomarker variable. Categorical and numeric clinical variables. No log transformation. Don't plot association between variables but plot biomarker and clinical variables.
```{r include=TRUE, fig.width=14, fig.height=10}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var=c("Country", "Age"), var.class=c("categorical", "numeric"),
             par.param = list(mfrow=c(2,2), cex.axis=1.2), show.association = FALSE,
             show.clinical.uni=TRUE)
```

## Categorial biomarker variable. Categorical and numeric clinical variables. No log transformation. Plot only association between variables.
```{r include=TRUE, fig.width=14, fig.height=6}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var=c("Country", "Age"), var.class=c("categorical", "numeric"),
             par.param = list(mfrow=c(1,2), cex.axis=1.2), 
             show.biomarker.uni = FALSE)
```

## Categorial biomarker variable. Categorical and numeric clinical variables. No log transformation. Plot only clinical variables.
```{r include=TRUE, fig.width=14, fig.height=6}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var=c("Country", "Age"), var.class=c("categorical", "numeric"),
             par.param = list(mfrow=c(1,2), cex.axis=1.1),
             show.biomarker.uni = FALSE, show.association = FALSE, 
             show.clinical.uni = TRUE)
```
## Categorial biomarker variable. Categorical and numeric clinical variables. No log transformation. Plot only biomarker variable.
```{r include=TRUE, fig.width=14, fig.height=6}
PlotProperty(data=input, biomarker.var="KRAS.mutant", biomarker.class="categorical",
             var=c("Country", "Age"), var.class=c("categorical", "numeric"),
             par.param = list(cex.axis=1.2),
             show.association = FALSE)
```


# PlotRspBar(): Compare response outcome across different population

## Plot response of one population

```{r, fig.width=4, fig.height=7}
PlotRspBar (input, outcome.var="Response",
            rsp.levels=c("CR", "PR","SD","NON CR/PD", "PD","NE"))
```

## Plot response of one population, binarize outcome (classify classes to responder vs non responder)

```{r, fig.width=4, fig.height=7}
PlotRspBar (input, outcome.var="Response", 
            binary=TRUE,
            rsp.response = c("CR","PR"),
            rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA))
```

## Change color

```{r, fig.width=4, fig.height=7}
PlotRspBar (input, outcome.var="Response", 
            binary=TRUE,
            rsp.response = c("CR","PR"),
            rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
            col=c("green","orange"))
```

## Plot horizontally
```{r, fig.width=7, fig.height=4}
PlotRspBar (input, outcome.var="Response", 
            binary=TRUE,
            rsp.response = c("CR","PR"),
            rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
            horiz=TRUE)
```

## By arm

```{r fig.width=7, fig.height=7}
PlotRspBar (input, outcome.var="Response", 
            binary=FALSE,
            rsp.levels=c("CR", "PR","SD","NON CR/PD", "PD","NE"),
            trt="Arm")

PlotRspBar (input, outcome.var="Response", 
            binary=TRUE,
            rsp.response = c("CR","PR"),
            rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
            trt="Arm")
```

## Compare full population vs BEP

```{r fig.width=7, fig.height=7}

PlotRspBar (input, outcome.var="Response", 
            binary=TRUE,
            rsp.response = c("CR","PR"),
            rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
            trt="Arm", compare.bep=TRUE,bep="BEP")


```

## Compare subgroups

```{r fig.width=7, fig.height=7}
PlotRspBar (input, outcome.var="Response", 
            binary=TRUE,
            rsp.response = c("CR","PR"),
            rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
            trt="Arm", compare.var=TRUE,var="KRAS.mutant")
```

## Plot count instead of percentage

```{r fig.width=7, fig.height=7}
PlotRspBar (input, outcome.var="Response", 
            binary=FALSE,
            rsp.levels=c("CR", "PR","SD","NON CR/PD", "PD","NE"),
            trt="Arm", 
            compare.var=TRUE,var="Sex", plot.count = TRUE)

```


# PlotKM(): Plot KM curves for subpopulations


## Without subgroup 

```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS ITT"))
```

## Without subgroup, BEP
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, bep="BEP",
             tte="PFS",cen="PFS.event", main="PFS BEP"))
```

## By TRT
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS ITT by treatment", trt="Arm"))
```

## By TRT, change color, line type
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS ITT by treatment", trt="Arm",
             col=c("orange","brown"),lty=c(1,2)))

```


## Mark median PFS, no grid
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS ITT by treatment", 
       trt="Arm", 
       plot.grid = FALSE, 
       plot.median=T))
```

## By KRAS.mutant only
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="OS BEP by treatment, by KRAS mutation", 
              var="KRAS.mutant"))
```

```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="OS BEP by treatment, by KRAS mutation", 
              var="KRAS.mutant", col=c("skyblue","darkgray")))
```

## By TRT and KRAS.mutant
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="OS BEP by treatment, by KRAS mutation", 
             trt="Arm", var="KRAS.mutant"))
```

## By TRT and KRAS.mutant, change color and line type 

```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS BEP by treatment, by KRAS mutation", 
             trt="Arm", var="KRAS.mutant",
             col=c("orange","orange","brown","brown"),
             lty=c(3,1,3,1)))
```

## By TRT and KRAS.expression (continuous biomarker, cut at median)

```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS BEP by treatment, by KRAS expression", 
             trt="Arm", var="KRAS.exprs", var.class="numeric",
             percentile.cutoff=0.5,xlim=c(0,18))
             )
```


## By TRT and KRAS.expression (continuous biomarker, cut at a numerical cutoff)

```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS BEP by treatment, by KRAS expression", 
             trt="Arm", var="KRAS.exprs", var.class="numeric",
             numerical.cutoff=100)
             )
```

## By TRT and KRAS.expression (continuous biomarker, cut at a numerical cutoff). High group defined as "greater than", instead of "greater than or equal to"


```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS BEP by treatment, by KRAS expression", 
             trt="Arm", var="KRAS.exprs", var.class="numeric",
             numerical.cutoff=100, equal.in.high = F)
             )
```

## By TRT and KRAS.expression , more than 2 groups


```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event",  
             main="PFS BEP by treatment, by KRAS expression", 
             trt="Arm", var="KRAS.exprs", var.class="numeric",
             numerical.cutoff=c(100,500), xlim=c(0, 20))
             )
```

```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             main="PFS BEP by treatment, by KRAS expression", 
             trt="Arm", var="KRAS.exprs", var.class="numeric",
             numerical.cutoff=c(100,500),
             col=c("green","green","green","brown","brown","brown"),
             xlim=c(0,20))
             )
```

## More flexibility, reorder subgroups

```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="OS",cen="OS.event", 
             main="OS BEP by treatment, by KRAS mutation", 
             varlist=c("Arm","KRAS.mutant"), 
             varlist.levels=list(c("TRT","CTRL"),c("Wild Type","Mutant")),
             legend.loc="left"))
```

## By TRT and KRAS.mutant, reorder and rename 
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="OS",cen="OS.event", bep="BEP",
             main="OS BEP by treatment, by KRAS mutation", varlist=c("Arm","KRAS.mutant"), 
             varlist.levels=list(c("TRT","CTRL"),c("Wild Type","Mutant")),
             varlist.labels=list(c("trt","ctrl"),c("wt","mut")),
               plot.median=T,legend.loc="left"))
```

## Legend location
```{r  include=TRUE,fig.width=7, fig.height=7}
print(PlotKM(data=sample.data, tte="PFS",cen="PFS.event", 
             var=c("Arm","KRAS.mutant"), 
             bep="BEP", legend.loc=NULL, legend.x=5, legend.y=.8))
```

# PlotTabForestBiomarker() : forest plot and summary statistics table for a single biomarker (include cutoff exploration for cont. biomarker )


## Survival outcome, 2-arm, categorical variable

by default BEP will be defined as patients non-NA biomarker measurement

```{r, fig.height=5,fig.width=7 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical")
```

## Survival outcome, another format


```{r, fig.height=10,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical",
                                  tabforest = TRUE)
```

## Survival outcome, 2-arm, categorical variable, don't show ITT, BEP, rename variable name in display

user can also define BEP column

```{r, fig.height=5,fig.width=7 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical",
                                  bep = 'BEP', 
                                  bep.indicator=1)
```

```{r, fig.height=4,fig.width=6 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical", 
                                  var.name="KRAS mut",
                                  show.itt=FALSE, 
                                  show.bep=FALSE)
```


## Survival outcome, 2-arm, continuous variable, greater than some percentage cutoffs

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=c(.25,.5,.75), 
                              #  cols=c("black","black","darkgreen","darkgreen","darkgreen"),
                                  numerical.cutoff=NULL,
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE)
```

## Survival outcome, 2-arm, continuous variable, less than some percentage cutoffs

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=c(.25,.5,.75), 
                                  numerical.cutoff=NULL,
                                  greater=FALSE, less=TRUE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE,
                                  bep = 'BEP', bep.indicator=1)
```

## Survival outcome, 2-arm, continuous variable, greater and less than some percentage cutoffs

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=c(.25,.5,.75), 
                                  numerical.cutoff=NULL,
                                  greater=TRUE, less=TRUE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE)
```

"Less" rows next to "Greater" rows:

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=c(.25,.5,.75), 
                                  numerical.cutoff=NULL,
                                  greater=TRUE, less=TRUE, greater.by.less = TRUE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE)
```

## Survival outcome, 2-arm, continuous variable, within cutoff bin

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=c(.25,.5,.75), 
                                  numerical.cutoff=NULL,
                                  greater=FALSE, less=FALSE,
                                  within.bin=TRUE,
                                  show.itt=TRUE, show.bep=TRUE)
```


## Survival outcome, 2-arm, continuous variable, greater than some numerical cutoffs

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=NULL,
                                  numerical.cutoff=c(50,100),
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE)
```

## Survival outcome, 2-arm, continuous variable, adjust for covariates

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=NULL,
                                  numerical.cutoff=c(50,100),
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE,
                                  bep = 'BEP', bep.indicator=1, covariate="Age")
```
## Survival outcome, 2-arm, continous variable, stratification
```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=NULL,
                                  numerical.cutoff=c(50,100),
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE,
                                  bep = 'BEP', bep.indicator=1, strata="Sex")
```
## Survival outcome, 2-arm, continuous variable, adjust for covariates and stratified by strata

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=NULL,
                                  numerical.cutoff=c(50,100),
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE,
                                  bep = 'BEP', bep.indicator=1,covariate="Age",strata="Sex")
```

## Survival outcome, 1arm 

within.bin, show.itt, show.bep will be ignored

```{r, fig.height=4,fig.width=6 }
PlotTabForestBiomarker(data=subset(input, Arm=="TRT"),
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt=NULL,
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=NULL,
                                  numerical.cutoff=c(50,100),
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE,covariate="Age",strata="Sex")
```

## Survival outcome, 2arm, flip TRT/CTRL order (calculate ctrl over trt HR)

```{r, fig.height=6,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=c(.25,.5,.75), 
                                  numerical.cutoff=NULL,
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE,
                                  covariate=NULL, #Sex
                                  strata=NULL, #Age
                                  placebo.code='TRT',
                                  active.code='CTRL')
```

## Survival outcome, 2-arm, continuous variable, greater than some percentage cutoffs, show across-arm and within-arm results in the same plot

```{r, fig.height=12,fig.width=9 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var="KRAS.exprs", 
                                  var.class="numeric", var.name="KRAS exprs",
                                  percentile.cutoff=c(.25,.5,.75), 
                                  numerical.cutoff=NULL,
                                  greater=TRUE, less=FALSE,
                                  within.bin=FALSE,
                                  show.itt=TRUE, show.bep=TRUE,
                       across.and.within = TRUE)
```



## Response outcome, 2-arm, categorical variable

by default BEP will be defined as patients non-NA biomarker measurement

```{r, fig.height=5,fig.width=7 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="binary",
                                  outcome.var=c("Response"),
                                  rsp.cat = TRUE,
                                  rsp.response = c("CR","PR"),
                                  rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical")
```

## Response outcome, another format


```{r, fig.height=10,fig.width=9 }
PlotTabForestBiomarker(data=input,                                  
                                  outcome.class="binary",
                                  outcome.var=c("Response"),
                                  rsp.cat = TRUE,
                                  rsp.response = c("CR","PR"),
                                  rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical",
                                  tabforest = TRUE)
```


## Continuous outcome, 2-arm, categorical variable


```{r, fig.height=5,fig.width=7 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="continuous",
                                  outcome.var=c("Lab_ontrt"),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical")
```

```{r, fig.height=10,fig.width=10 }
PlotTabForestBiomarker(data=input,
                                  outcome.class="continuous",
                                  outcome.var=c("Lab_ontrt"),
                                  trt="Arm",
                                  var="KRAS.mutant", 
                                  var.class="categorical",tabforest=T)
```



# PlotTabForestMulti(): Forest plots to compare multiple variates' effect in ITT vs BEP, or compare multiple variates' effect in subpopulations 

## 2-arm, compare BEP vs ITT


```{r, fig.height=12,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1,
                   compare.bep.itt=TRUE
                   )
```


## 2-arm, compare subgroups defined by biomarker 


```{r, fig.height=6,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1, compare.bep.itt=FALSE,
                   compare.subgroup=TRUE, subgroup="KRAS.mutant"
                   )
```


## 2-arm, compare subgroups defined by biomarker, also show ITT 

```{r, fig.height=6,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1, compare.bep.itt=FALSE,
                   compare.subgroup=TRUE, subgroup="KRAS.mutant", show.itt=TRUE
                   )
```

## 2-arm, compare subgroups defined by biomarker, also show BEP

```{r, fig.height=12,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1, compare.bep.itt=FALSE,
                   compare.subgroup=TRUE, subgroup="KRAS.mutant", show.itt=TRUE, show.bep=TRUE
                   )
```

If bep column is not defined, the program will take the non NA entries in subgroups column as BEP

```{r, fig.height=12,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var=c("Sex","Age"), compare.bep.itt=FALSE,
                   compare.subgroup=TRUE, subgroup="KRAS.mutant", show.itt=TRUE, show.bep=TRUE
                   )
```


## 1-arm, compare ITT vs BEP


```{r, fig.height=6,fig.width=9 }
PlotTabForestMulti(data=subset(input,Arm=="TRT"),
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt=NULL,
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1, compare.bep.itt=TRUE
                   )
```

## 2-arm, compare subgroups defined by biomarker, multiple cutoffs for continuous biomarker 


```{r, fig.height=12,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm", percentile.cutoff=c(.33,.66),
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1, compare.bep.itt=FALSE,
                   compare.subgroup=TRUE, subgroup="KRAS.mutant"
                   )
```

## 2-arm, compare subgroups defined by biomarker, multiple cutoffs for continuous biomarker, calculate "< cutoff" 


```{r, fig.height=12,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm", percentile.cutoff=c(.33,.66),less=TRUE,greater=FALSE,
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1, compare.bep.itt=FALSE,
                   compare.subgroup=TRUE, subgroup="KRAS.mutant"
                   )
```

## 2-arm, compare subgroups defined by biomarker, multiple cutoffs for continuous biomarker, calculate within bin statistics 


```{r, fig.height=8,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm", percentile.cutoff=c(.33,.66),within.bin=TRUE,
                                  var=c("Sex","Age"),
                   bep="BEP",bep.indicator=1, compare.bep.itt=FALSE,
                   compare.subgroup=TRUE, subgroup="KRAS.mutant"
                   )
```

## Basic forest plot without comparison

```{r, fig.height=6,fig.width=9 }
PlotTabForestMulti(data=input,
                                  outcome.class="survival",
                                  outcome.var=c("PFS","PFS.event"),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   compare.bep.itt=FALSE, compare.subgroup=FALSE,itt.name=""
                   )
```

## Response outcome


```{r, fig.height=6,fig.width=9 }
PlotTabForestMulti(data=input,
                                                                   outcome.class="binary",
                                  outcome.var=c("Response"),
                                  rsp.cat = TRUE,
                                  rsp.response = c("CR","PR"),
                                  rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   compare.bep.itt=FALSE, compare.subgroup=FALSE,itt.name=""
                   )
```


```{r, fig.height=10,fig.width=10 }
PlotTabForestMulti(data=input,
                                                                   outcome.class="binary",
                                  outcome.var=c("Response"),
                                  rsp.cat = TRUE,
                                  rsp.response = c("CR","PR"),
                                  rsp.nonresponse = c("SD", "PD","NON CR/PD","NE",NA),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   compare.bep.itt=FALSE, compare.subgroup=FALSE,itt.name="", tabforest=T
                   )
```

## Continuous endpoint


```{r, fig.height=6,fig.width=9 }
PlotTabForestMulti(data=input,
                               outcome.class="continuous",
                                  outcome.var=c("Lab_ontrt"),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   compare.bep.itt=FALSE, compare.subgroup=FALSE,itt.name=""
                   )
```


```{r, fig.height=10,fig.width=10 }
PlotTabForestMulti(data=input,
                 outcome.class="continuous",
                                  outcome.var=c("Lab_ontrt"),
                                  trt="Arm",
                                  var=c("Sex","Age"),
                   compare.bep.itt=FALSE, compare.subgroup=FALSE,itt.name="", tabforest=T
                   )
```


# PlotSTEPP(): STEPP (Subpopulation Treatment Effect Pattern Plot) from the given point estimates and confidence intervals at desired percentiles.

## Survival outcome
```{r}
PlotSTEPP(data = input,
          outcome.var = c("PFS", "PFS.event"),
          outcome.class = "survival",
          trt = "Arm",
          var = "KRAS.exprs",
          placebo.code = "CTRL",
          active.code = "TRT",
          csv.name = NULL,
          pdf.name = NULL
) 
```

## Continuous outcome
```{r}
PlotSTEPP(data = input,
          outcome.var = "Baseline.SLD",
          outcome.class = "continuous",
          trt = "Arm",
          var = "KRAS.exprs",
          covariate= "Sex",
          placebo.code = "CTRL",
          active.code = "TRT",
          csv.name = NULL,
          pdf.name = NULL
)
```

## Binary outcome variable 
```{r}
PlotSTEPP(data = input,
          outcome.var = "ECOG",
          outcome.class = "binary",
          trt = "Arm",
          var = "KRAS.exprs",
          placebo.code = "CTRL",
          active.code = "TRT",
          csv.name = NULL,
          pdf.name = NULL
)
```


# CoxTab(): fit cox proportional model for multiple covariates (additive model or separate models)

## single covariate

```{r}
CoxTab(data=sample.data, tte="OS", cens="OS.event",bep='BEP', var='Sex' )


```

## multiple covariates, ITT
Additive model will be fitted.
```{r}
kable(
CoxTab(data=sample.data,tte="OS", cens="OS.event",  var=c('Sex',"Country","Age"), 
       var.class=c("categorical","categorical","numeric"))
)
```

Simplify: automatically learn var.class from column class:

```{r}
kable(
CoxTab(data=sample.data,tte="OS", cens="OS.event",  var=c('Sex',"Country","Age"))
)

```

## BEP

```{r}
kable(
CoxTab(data=sample.data,tte="OS", cens="OS.event",  var=c('Sex',"Country","Age"), 
       var.class=c("categorical","categorical","numeric"), bep="BEP")
)
```


## Reorder Sex: M as reference

```{r}
kable(
CoxTab(data=sample.data,tte="OS", cens="OS.event",  var=c('Sex',"Country","Age"), 
       var.class=c("ordered.factor","categorical","numeric"), 
       ordered.factor.levels.list=list(Sex=c("M","F")),bep="BEP")
)
```

## Fit separate models for each variable instead of fitting an additive model
```{r}
kable(
CoxTab(data=sample.data,tte="OS", cens="OS.event",  var=c('Sex',"Country","Age"),
       additive=FALSE)
)

```

# SummaryTwoGroups()

## Contunuous outcome
###  Create summary statistics
```{r}
SummaryTwoGroups(outcome.var = input$OS, treatment.var = input$Arm, 
            placebo.code = "CTRL", active.code = "TRT", 
            outcome.class = "continuous")
```
### Add a covariate variable
```{r}
SummaryTwoGroups(outcome.var = input$OS, treatment.var = input$Arm, 
            placebo.code = "CTRL", active.code = "TRT", 
            outcome.class = "continuous", covariate.var = input$Sex)
```
### Add return.fit = TRUE to return a table of summary statistics
```{r}
SummaryTwoGroups(outcome.var = input$OS, treatment.var = input$Arm, 
            placebo.code = "CTRL", active.code = "TRT", 
            outcome.class = "continuous", covariate.var = input$Sex, 
            return.fit = TRUE)
```

##  Binary outcome
### Create summary statistics
```{r}
SummaryTwoGroups(outcome.var = input$OS.event, treatment.var = input$Arm, 
            placebo.code = "CTRL", active.code = "TRT", 
            outcome.class = "binary")
```

## Survival variable 
### Create summary statistics
```{r}
SummaryTwoGroups(outcome.var = input[, c("OS", "OS.event")], 
            treatment.var = input$Arm, placebo.code = "CTRL", 
            active.code = "TRT", outcome.class = "survival")
```
### Add a covariate variable
```{r}
SummaryTwoGroups(outcome.var = input[, c("OS", "OS.event")], 
            treatment.var = input$Arm, placebo.code = "CTRL", 
            active.code = "TRT", outcome.class = "survival", 
            covariate.var = input$Sex)
```
### Add a stratification variable
```{r}
SummaryTwoGroups(outcome.var = input[, c("OS", "OS.event")], 
            treatment.var = input$Arm, placebo.code = "CTRL", 
            active.code = "TRT", outcome.class = "survival", 
            covariate.var = input$Sex, strat.factor.var = input$Age)
```

# LogRankTab(): log rank test for subgroup analysis

```{r}
  kable(
    LogRankTab(data=input,tte="PFS",cens="PFS.event",var="Arm")
  )
```

# BoxPlot(): advanced box plot function


## Generate example dataset
```{r}
example <- data.frame( y=c(rnorm(30)+10, rnorm(4)+20, rnorm(15)+15, NA), 
 							time=c(rep("t2", 30), rep("t4",4), rep("t1", 15), "t3"),
 							grp=sample(1:3, 50, TRUE), sex=sample(1:2, 50, TRUE))
head(example)
str(example)
```

## Specify data as named list
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(list(a=rnorm(50,,3), b=rnorm(25,1,4), c=rnorm(75,2,1)))
```

## Same plot, now horizontally plotted
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(list(a=rnorm(50,,3), b=rnorm(25,1,4), c=rnorm(75,2,1)), horizontal=TRUE,
 			 Xaxis=list(las=2, hadj=2), Xaxis2=list(las=2, hadj=-.25))
```

## Specify data as numeric matrix
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(matrix(rnorm(100), ncol=4, dimnames=list(NULL, LETTERS[1:4])) )
```
 
## Specify data as numeric (unnamed) vectors
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(rnorm(50,,3), rnorm(25,1,2), rnorm(75,2,1))
```
 
## ... horizontally plotted (no axis-label rotation done here)
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(rnorm(50,,3), rnorm(25,1,2), rnorm(75,2,1), horizontal=TRUE)
```

## Plot values 'y' according to time 'time' (factor levels are automatically ordered as e.g. function sort does)
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example, y~time, sc.pch=16)
```
 
## Now as box-percentile plot
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example, y~time, sc.pch=16, box.type="bp")
```
 
## With custom main title
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example, y~time, sc.pch=16, box.type="bp", Title=list(main="Custom Main Title"))
```

## The same plot with trend line (connects either means or medians) and Y-axis label
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example, y~time, trend="median", ylab="Y-Axis Label")
```

## Use an addition grouping variable to color points in the stripchart
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example, y~time, trend="median", ylab="Y-Axis Label",
           sc.col=c("red", "blue", "green")[example$grp] )
``` 

## Use yet another grouping factor for plotting symbols in the stripchart
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example, y~time, trend="median", ylab="Y-Axis Label",
           sc.col=c("red", "blue", "green")[example$grp],
 		     sc.pch=c(5,10)[example$sex] )
 
legend("bottomright", fill=c("red", "green", "blue", "white", "white"), 
           legend=c("Stage I", "Stage II", "Stage III", "Female", "Male"), 
           pch=c(-1, -1, -1, 5, 10), border=NA)
```

## Generate new dataset with different stucture
```{r include=TRUE,fig.width=7, fig.height=4}
example2 <- data.frame(y1=12+rnorm(15), y2=15+rnorm(15), y3=25+rnorm(15))
```
 
## Boxplot with trend lines (mean) and grid added to the plot
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example2, var=c("y1", "y2", "y3"), Grid=TRUE, trend="mean")
```

## Now as box-percentile plot
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(example2, var=c("y1", "y2", "y3"), box.type="bp", Grid=TRUE)
```

## Use yet another way to specify the data
```{r include=TRUE,fig.width=7, fig.height=4}

BoxPlot(rnorm(1856, 5), runif(1245, 2,10), exp(rnorm(2311)), sc.col=as.rgb("black", .05), 
            box.type="bp", Xaxis=list(labels=c("~N(5,1)", "~Unif(2,10)", "~exp(N(0,1))")), ylim=c(0,15))     
```
 
## Specifying data as numeric matrix and using a trend-line connecting the means
```{r include=TRUE,fig.width=7, fig.height=4}
    mat <- matrix(c(rep(10,45), rep(15,45), rep(35,45))+rnorm(135), ncol=3)
    BoxPlot(mat, trend="mean", trend.col="red", Ylabel=list(text="Example Measurements"))
```

## Multiple grouping factors can be specified via the formula interface which is exemplified using the mtcars dataset, of interest is miles per gallon (mgp) depending on number of gears and on the number of cylinders
```{r include=TRUE,fig.width=7, fig.height=4}
data(mtcars)
BoxPlot(mtcars, mpg~gear:cyl)  
``` 
 
## Now 'cyl' is explicitly nested within 'gear' which changes the ordering of combined grouping factors (which is identical to using formula 'mpg~cyl:gear').
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(mtcars, mpg~cyl %in% gear)     
```

## More meaningful group-labels are best specified using custom factor-level names (increase width of the plot window)
```{r include=TRUE,fig.width=7, fig.height=4}
dat <- mtcars
dat$cyl <- factor(dat$cyl, levels=c(4,6,8), labels=c("4Cyl", "6Cyl", "8Cyl"))
dat$gear <- factor(dat$gear, levels=c(3,4,5), labels=c("3Gear", "4Gear", "5Gear")) 
BoxPlot(dat, mpg~cyl %in% gear)
```

## One can use a table as Xaxis label representing the factor-level combination defining sub-classes 
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(dat, mpg~cyl:gear, XaxisTab=list(),mar=c(8,3,5,1))
```
 
## With the original factor levels and as horizontal plot
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(mtcars, mpg~cyl:gear, XaxisTab=list(), mar=c(5,8,5,4), 
 			 horizontal=TRUE, Ylabel=list(text="Y-axis label now appearing on X-axis"))
```

## Using smaller bottom margin will result in smaller table height
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(dat, mpg~cyl:gear, XaxisTab=list(font=2, col="darkblue", cex=1.25), mar=c(5,3,5,1))
```
 
## One can use different font-settings for rownames and cells of the table
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(dat, mpg~cyl:gear, XaxisTab=list(Label=list(font=2, col="darkblue", cex=1.25), 
 			 Text=list(col="red")), mar=c(5,3,5,1))
```

## Use more crossed factors
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(dat, mpg~cyl:gear:vs, XaxisTab=list(), mar=c(5,3,5,1))
```

## Alternatively one can use the 'Xaxis' argument, but the ordering of these labels is not checked which is not important for automatically generated group-labels as shown in the previous example
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(mtcars, mpg~cyl %in% gear, 
 			 Xaxis=list(labels=paste(rep(c("4Cyl", "6Cyl", "8Cyl"),3), 
 						c(rep("3Gear",3), rep("4Gear",3), rep("5Gear",3)), sep=".")))
```

## The same plot with some fancy options
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(dat, mpg~cyl %in% gear, Title=list(main="Miles per Gallon by Number of Gears", 
 			 col.main="Green", cex.main=2.5), vline=c(3.5, 6.5), vl.lty=2, vl.col="gray", vl.lwd=2,
           Xaxis=list(labels=NA, at=1:9, tick=TRUE), col=c(rep("blue", 3), rep("red", 3), rep("green", 3)),
           Xaxis2=list(tick=FALSE), Yaxis=list(at=seq(10,34,2)), Grid=list(x=1:9, y=seq(10,34,2)),
           Xlabel=list(text=paste(rep(c("4Cyl", "6Cyl", "8Cyl"),3), c(rep("3Gear",3), rep("4Gear",3), rep("5Gear",3))),
                       at=1:9, las=2, adj=1, line=0.75, col=c(rep("blue", 3), rep("red", 3), rep("green", 3))), 
           mean.col=c(rep("cyan", 3), rep("orange", 3), rep("magenta", 3)), Box=FALSE, trend="mean",mar=c(6,4,5,2) )
```
 
## Horizontal fancy plot  
```{r include=TRUE,fig.width=7, fig.height=4}
BoxPlot(mtcars, mpg~cyl %in% gear, Title=list(main="Miles per Gallon by Number of Gears", col.main="#84d52b", cex.main=1.5), 
 		vline=c(3.5, 6.5), vl.lty=2, vl.col="gray", vl.lwd=2,
 		Xaxis2=list(tick=FALSE, las=2, hadj=-.25), Yaxis=list(at=seq(10,34,2)), Grid=list(x=1:9, y=seq(10,34,2)),
 		mean.col=c(rep("cyan", 3), rep("orange", 3), rep("magenta", 3)), Box=FALSE, trend="mean",
 		mar=c(3, 7, 4, 4), horizontal=TRUE, sc.pch=c(0, 15)[dat$am+1], sc.col="wheat4",
 		XaxisTab=list(Text=list(col=c(rep("cyan", 3), rep("orange", 3), rep("magenta", 3)))) )
 
 legend(x="topright", pch=c(0, 15), legend=c("automatic", "manual"), box.lty=0, col="wheat4")
```

# Longitudinal analysis

In the context of biomarker analysis, the `PlotLong` function can be used to plot the longitudinal measurement of a biomarker endpoint for one or more patient subpopulations. Here we'll use the packaged dataset `bmkr` (biomarker) which has columns for patient ID (`pid`), treatment (`trt`), gender (`sex`), visitation month (`vm`) and endpoint reading (`ep`)

## Biomarker Endpoint Progression

### Mean +/- SEM Timecourse

Here we plot the mean and standard error of the mean over time. We can see clearly that the mean biomarker endpoint increases throughout the study.

```{r}
PlotLong(longbmkr, aes(x=vm, y=ep),
         xlab = 'Visitation Month',
         ylab = 'Biomarker Endpoint',
         labs.title = 'Mean +/- SEM')
```

### Tukey Hinges and Whiskers

However, we may be more interested in the population distrubtion progression where we'd want to plot the Tukey boxplot hinges and whiskers over time. From this plot, we can see that although there is a slow rise in population mean, the patient response also becomes more distributed. 

```{r}
PlotLong(longbmkr, aes(x=vm, y=ep), fun.data = 'tukey',
         xlab = 'Visitation Month',
         ylab = 'Biomarker Endpoint',
         labs.title = 'Tukey Whiskers and Hinges')
```

## Plotting Subpopulations

```{r}
PlotLong(longbmkr, aes(x=vm, y=ep, group=trt, color=trt, fill=trt), 
         fun.data = 'tukey', facet.fun = . ~ sex,
         xlab = 'Visitation Month',
         ylab = 'Biomarker Endpoint',
         labs.title = 'Biomarker Timecourse',
         labs.caption = 'Ribbons represent Tukey hinges and whiskers')
```

## Including sample counts

### Sample counts as table above plot

```{r, fig.height = 3.5,eval=TRUE}
library(dplyr)
PlotLong(longbmkr %>% filter(vm <= 24), 
         aes(x=vm, y=ep, group=trt, color=trt, fill=trt), 
         fun.data = 'tukey', facet.fun = . ~ sex,
         show.counts = 'table',
         xlab = 'Visitation Month',
         ylab = 'Biomarker Endpoint',
         labs.title = 'Biomarker Timecourse',
         labs.caption = 'Ribbons represent Tukey hinges and whiskers')
```

## Plotting Deviations as Errorbars

```{r, fig.height = 3.5, eval=TRUE}
PlotLong(longbmkr %>% filter(vm <= 24), 
         aes(x=vm, y=ep, group=trt, color=trt, fill=trt), 
         fun.data = 'tukey', facet.fun = . ~ sex,
         show.counts = 'table',
         plot.style = 'errorbars',
         xlab = 'Visitation Month',
         ylab = 'Biomarker Endpoint',
         labs.title = 'Biomarker Timecourse',
         labs.caption = 'Ribbons represent Tukey hinges and whiskers')
```


